---
title: "Cyclone Landfall Frequency Analysis: <br> Identifying High-Risk Countries and Territories on the American continent"
abstract: "Using Bayesian methods and cylone data from 2000 to 2024, this report tracks and forecasts tropical cyclone landfall occurrences on the American continent. Seasonality patterns and yearly trends are investigated as well as regions most vulnerable to tropical cyclone impacts. The analysis showed a clear seasonal pattern, with most cyclone landfalls expected between June and November, yet no significant yearly trends, suggesting there is no strong long-term change in landfall occurrences over the studied period. The United States of America and Mexico stand out as the most high risk countries, with 30 and 28 predicted landfalls, respectively,  over the next five years."
output: pdf_document
bibliography: ["references.bib"]
csl: harvard-imperial-college-london.csl
link-citations: true
---

## 1. Introduction

Tropical cyclones are among the most destructive natural disasters, causing the highest number of casualties compared to all other natural disasters [@casualties] as well as severe economic losses globally. These storms cause the most damage as they move toward land, bringing along extreme winds, heavy rainfall, storm surges and floods [@landfall-damages]. Even when coastal communities are evacuated, cyclones can still cause extreme damage to properties and infrastructure as well as a significant loss of agricultural production. In addition to these, cyclone landfalls have long-term impacts, including economic slowdown, loss of coastal ecosystems and wildlife habitats, loss of businesses (especially local/family-owned enterprises), permanent displacement of coastal communities and significant rebuilding costs that can strain government budgets. More frequent and repeated exposure to these landfalls limits recovery time, leaving areas unprotected or poorly rebuilt, further exacerbating the associated damages.

Governments and disaster relief organisations have put together a variety of measures over the years, intended to mitigate these risks, including early warning systems, evacuation plans and strengthening infrastructure. However, the unpredictability of cyclone landfalls and the extent of regional disparities makes it crucial to understand landfall patterns and trends, in order to optimise response strategies tailored to communities in need.

Historically, landfall locations and frequency have been quite variable, but certain countries, such as the United States of America and Mexico, experience higher incidences than others. As such, understanding and accurately forecasting when and where cyclones are most likely to strike is essential for assessing the vulnerability of different areas, resource allocation and disaster planning. By identifying the areas at greatest risk as well as the landfall seasonal peaks, governments and organisations can ensure that resources such as emergency supplies, medical teams, and infrastructure support are on hand.

In this report, we focus on the evolution of hurricane landfall frequency since 2000 as well as analysing any monthly trends, by using Bayesian Poisson regression models and cyclone track data. We restrict our analysis to the American continent, investigating landfalls originating in the Atlantic basin (Atlantic Ocean, Caribbean Sea, and Gulf of America) and the Eastern Pacific basin (which extends from the western coast of Mexico and Central America to 140°W) [@eastern-pacific]. Our aim is to identify monthly patterns and long-term trends in landfall frequency, as well as to highlight the most vulnerable countries and territories, in order to guide resource allocation and optimise disaster preparedness and response.


## 2. Data and Methodology

We based our analysis on the Atlantic hurricane database (HURDAT2), 1851-2024, more recently updated on April 4th, 2025 to include in the 2024 hurricane season, as well as the Northeast and North Central Pacific hurricane database (HURDAT2), 1949-2024, most recently updated on March 17, 2025 to include the 2024 hurricane season. These data sets are provided by the National Hurricane Center (NHC), as part of the National Oceanic and Atmospheric Administration (NOAA) and provides records of cyclone coordinates, maximum winds, central pressure, system status at 6 hour intervals. The HURDAT2 databases also include additional records outside of the standard intervals if the cyclone makes landfall, unexpectedly changes status or intensity or reaches a peak in terms of wind speed or pressure. 

Although the Atlantic HURDAT2 database goes back to 1851, the NHC indicates that track data from the nineteenth century is often incomplete and inaccurate due to less advanced technology and therefore under reported and under analysed measurements. Moreover, international landfalls are only marked from 1951 to 1970 and 1991 onward, leading to significant gap in historical data. 

Our analysis therefore focuses on landfall trends from 2000 to 2024, thus ensuring consistency and making use of improved landfall measurements. However, it is important to note that this introduces some limitations to our models, such as evaluating long term trends using a relatively short time period and  sparse data, making our model predictions more susceptible to our initial assumptions.

Once we collected our data, we used geographical data, containing country outlines and coordinates, and combined these with our landfall coordinates to determine where the cyclone crossed a coastline. We say a cyclone makes landfall when the center of the system crosses a coastline. This center, however, usually ranges from 32 to 64km and therefore, certain landfall coordinates did not point to any land. In these cases, we mapped the landfall data points to the nearest country. We also note that certain cyclones make more than one landfall, however for the purpose of this report, our analysis focuses solely on the number of landfalls.

We first modeled the number of cyclone landfalls per month and year using Bayesian Poisson regression models [@Elsner], in order to explore any seasonality effects and yearly trends. Our first model included both a monthly effect and a non-linear yearly effect, due to the high fluctuations in cyclone landfall numbers per year. Our second model only included the monthly effect, ignoring any long-term trends. We then used a statistical method called Leave-One-Out Cross Validation (LOO) to select the model with the best performance, in this case, our second model. 

Our third model then investigated where cyclones were most likely to land, amongst the top 10 countries and commonwealths which have experienced the most cyclone landfalls since 2000. These include Antigua and Barbuda, the Bahamas, Belize, Canada, Cuba, the Dominican Republic, Mexico, Nicaragua, Puerto Rico and the United States of America.

Using our best performing seasonal model and our geographical model, we then simulated landfall counts for each month as well as for each of the ten most affected countries and territories over the next five years (2025 to 2029). These forecasts thus support preparedness, planning and resource allocation in high risk areas.

For more details on methodology and technical analysis, please refer to the Appendix.

## 3. Results

### 3.1 Cyclone trends over time

- plots of yearly number of cyclones and monthly cyclone numbers/ landfall frequency

- talk about models (again not very technical)

- landfall probability plots

- landfall predictions

- what this means

### 3.2 areas at high risk

- map plots

- talk about model

- landfall probabiltiy plots

- landfall predictions

- what this means


## 4. Recommendations

## 5. Limitations of models and data biases

## 6. Conclusion

## References

<div id="refs"></div>

## Appendix - Technical description of analysis and data processing


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
```




```{r, echo=FALSE, fig.align='center'}
pdf_file <- here("outputs", "eda-hurricane-data", "landfall-map.pdf")
knitr::include_graphics(pdf_file)
```